---
title: "MLAI 2025 Regression Challenge"
output:
  pdf_document: default
  html_document:
    includes:
      before_body: Title.tex
---

# Project: MLAI 2025 Regression Challenge

**Team Members**: Gabriel Witte, Nataliia Korobkova, Muhammad Saim, Ali Areeb, Alina Porschke

**Description**: Here we present a detailed ML pipeline following the CRISP-DM standard for the AirBnB dataset.

## Step 0. Imports and dependencies

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# care for nessesary packages
required_pkgs <- c("ranger", "kernlab", "glmnet", "geosphere")
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(ggplot2)
library(tidymodels)
library(dplyr)
library(kernlab)
library(ranger)
library(glmnet)
library(geosphere)
library(tidyverse)
library(tidymodels)
library(stringr)
library(lubridate)
library(janitor)
library(tidytext)
library(stopwords)
library(scales)
library(corrplot)
library(dplyr)
library(xgboost)
library(Matrix)
library(yardstick)
library(stringr)
library(stringi)
library(vip)

set.seed(42) # we set a deterministic seed to output the same results no matter how we train the model

# we use this to remove outliers later. 
remove_outliers_iqr <- function(x) {
  if (!is.numeric(x)) stop("Input must be numeric")

  Q05 <- quantile(x, 0.05, na.rm = TRUE)
  Q95 <- quantile(x, 0.95, na.rm = TRUE)
  IQR_value <- IQR(x, na.rm = TRUE)

  lower_bound <- Q05 - 1.5 * IQR_value
  upper_bound <- Q95 + 1.5 * IQR_value

  x >= lower_bound & x <= upper_bound
}

make_perf_table <- function(res, wf, train_df, truth, metric_name = "rmse") {

  val_raw <- collect_metrics(res) %>% dplyr::filter(.metric == metric_name)

  # Identify parameter vs meta columns
  meta_candidates <- c(".metric", ".estimator", "mean", "std_err",
                       ".config", ".iter", ".sample", "n", ".n")
  meta_cols  <- intersect(meta_candidates, names(val_raw))
  param_cols <- setdiff(names(val_raw), meta_cols)

  # Figure out which name is used for resample count
  n_resample_col <- dplyr::case_when(
    "n"  %in% names(val_raw) ~ "n",
    ".n" %in% names(val_raw) ~ ".n",
    TRUE ~ NA_character_
  )

  val_tbl <- val_raw %>%
    dplyr::transmute(
      !!!rlang::syms(param_cols),
      rmse_val = mean,
      se_val   = std_err,
      n_resamples = if (!is.na(n_resample_col)) .data[[n_resample_col]] else NA_integer_
    )

  # Fit on training data for each param row
  params_unique <- dplyr::distinct(val_tbl, !!!rlang::syms(param_cols))

  train_tbl <- purrr::pmap_dfr(params_unique, function(...) {
    par_row <- tibble::tibble(...)
    wf_i    <- tune::finalize_workflow(wf, par_row)
    fit_i   <- parsnip::fit(wf_i, data = train_df)
    preds   <- predict(fit_i, new_data = train_df)[[".pred"]]

    tibble::as_tibble(par_row) %>%
      dplyr::mutate(
        rmse_train = yardstick::rmse_vec(truth = train_df[[truth]], estimate = preds)
      )
  })

  val_tbl %>%
    dplyr::left_join(train_tbl, by = param_cols) %>%
    dplyr::mutate(gap = rmse_val - rmse_train) %>%
    dplyr::arrange(rmse_val)
}
```

## Step: 1. Data Understanding

In this section we'll get a feel for the current variables we will work with.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train = read.csv("data/train.csv")

df_train %>%
  summarise(
    mean_price   = mean(price),
    median_price = median(price),
    sd_price     = sd(price),
    min_price    = min(price),
    max_price    = max(price)
  )

df_train %>%
 summary()

# In general we see the the following data:

# Variable: ID
# Data type: int
# Variable type: Numerical/Quantitative -> Discrete
# Description: Simple identifier for the rows.
# Useful for analysis: False, as it is mainly an identifier
# values: range(0, inf)

# Variable: name
# Data type: chr
# Variable type: Categorical/Qualitative -> Nominal
# Description: The name of the place being offered at AirBnB
# Useful for inference: False, as it's mainly an identifier
# values: any 

# Variable: host_id
# Data type: int
# Variable type: Numerical/Quantitative -> Discrete
# Description: Simple identifier - Won't be useful for the data analysis
# Useful for inference: False, as it's mainly an identifier
# values: range(0, inf)

# Variable: host_name
# Data type: chr
# Variable type: Categorical/Qualitative -> Nominal
# Description: The name of the host/franchise offering the place.
# Useful for inference: False, some names would be interesting to analyze, e.g companies such as ComfortInn, Ibis, etc. However AirBnB allows for people to host their hotels, meaning that names such as Rebecca may refer to multiple people, not just one.
# values: any

# Variable: neighbourhood
# Data type: chr
# Variable type: Categorical/Qualitative -> Nominal
# Description: The location in which the offering is based
# Useful for inference: Possible, have to check (will have to encode if so).
# values: shown below
unique(df_train$neighbourhood)

# Variable: latitude
# Data type: dbl
# Variable type: Numerical/Quantitative -> Continuous
# Description: The location in coordinates of the offering in the x-axis.
# Useful for inference: Possible, have to check.
# values: range(-180, 180)

# Variable: longitude
# Data type: dbl
# Variable type: Numerical/Quantitative -> Continuous
# Description: The location in coordinates of the offering in the y-axis.
# Useful for inference: Possible, have to check.
# values: range(-180, 180)

# Variable: room_type
# Data type: chr
# Variable type: Categorical/Qualitative -> Nominal
# Description: The type of room types that the offering provides.
# Useful for inference: Possible, have to check (will have to encode if so).
# values: shown below
unique(df_train$room_type)

# Variable: price
# Data type: dbl
# Variable type: Numerical/Quantitative -> Continuous
# Description: The price for the offering.
# Useful for inference: This is the TARGET.
# values: range(0, inf)

# Variable: minimum_nights
# Data type: int
# Variable type: Numerical/Quantitative -> Discrete
# Description: The minimum amount of nights a user must select in order to book the offering.
# Useful for inference: Possible, have to check.
# values: range(0, inf)

# Variable: number_of_reviews
# Data type: int
# Variable type: Numerical/Quantitative -> Discrete
# Description: The amount of reviews currently for the offer.
# Useful for inference: Possible, have to check.
# values: range(0, inf)

# Variable: last_review
# Data type: chr
# Variable type: Categorial/Qualitative -> Ordinal
# Description: The date of the last review.
# Useful for inference: Possible, have to check (will have to encode).
# values: anything in the format YYYY-MM-DD

# Variable: reviews_per_month
# Data type: dbl
# Variable type: Nominal/Quantitative -> Continuous
# Description: The amount of reviews the offering gets per month (likely average)
# Useful for inference: Possible, have to check.
# values: range(0.0 , inf)

# Variable: calculated_host_listings_count 
# Data type: int
# Variable type: Nominal/Quantitative -> Discrete
# Description: The amount of listings the host offers.
# Useful for inference: Possible, have to check.
# values: range(0, inf)

# Variable: availability_365
# Data type: int
# Variable type: Nominal/Quantitative -> Discrete
# Description: How many days a year the offering is available for booking
# Useful for inference: Possible, have to check.
# values: range(0, inf)
```

### 1.1 - Dropping redundant rows + setting correct data type

Based on the previous step, we found out that `ID`, `name`, `host_id`, and `host_name` are all just identifiers -\> therefore we simply remove them from out dataset. Also since `last_review` is a datetime, we change it to the date format:

however, we can still extract some keywords from `name` in order to maybe get more important information, so we keep it for now

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train <- df_train %>% 
  select(
    -ID, # dropping ID
    -host_id, # host_id 
    -host_name # host_name
  ) %>%
  mutate(
    last_review = ymd(last_review),
    neighbourhood = as.factor(neighbourhood)
  )

```

### 1.2 - Detecting Logical Inconsistencies and Impossible Values

Also based on the previous step, we can proceed to to check for illogical inconsistencies in the variables and between similar variables

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# the script checks our flag conditions and there outputs the columns which violate at least one of the conditions + outputs only columns where at least one record has violated it

df_issues <- df_train %>%
  mutate(
    flag_latitude = latitude < -90 | latitude > 90, #checking the longitude and latitude 
    flag_longitude = longitude < -180 | longitude > 180,
    flag_price = price <= 0, # price should be a positive number + a price of 0 is pretty sus
    flag_min_nights = minimum_nights < 1, # minimum nights also should be at least 1
    flag_availability = availability_365 < 0 | availability_365 > 365, # the bookable days in a year must not be negative and not more than 365 days 
    flag_host_listings = calculated_host_listings_count < 1, # a host must have at least one listing
    flag_reviews_neg = number_of_reviews < 0, #number of reviews cannot be negative
    flag_reviews_per_month_neg = reviews_per_month < 0, # same for reviews
    flag_review_logic = (number_of_reviews == 0 & reviews_per_month > 0), #if there are no reviews, reviews per month must be 0 (or N/A)
    flag_last_review_future = ymd(last_review) > Sys.Date(), #last review date cant be in the future
    flag_infinite = if_any(where(is.numeric), is.infinite) # here we just check whether there are infinite values
  ) %>%
  # here we check if at least one flag is true
  filter(if_any(starts_with("flag_"), ~ .x %in% TRUE)) %>%
  # here we get the columns that have been violated 
  {
    flag_cols <- grep("^flag_", names(.), value = TRUE)
    flags_to_keep <- names(
      select(., all_of(flag_cols)) %>%
        select(where(~ any(.x, na.rm = TRUE)))
    )
    non_flag_cols <- setdiff(names(.), flag_cols)
    select(., all_of(flags_to_keep), all_of(non_flag_cols))
  }

df_issues


```

the rows luckily do not violate any of the conditions, hence we can expect that most of the dataset is integrous (as in integrity)

### 1.3 - Dealing with missing values

now we deal with the missing values

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# to prettify, we print the missing count and percentage inside the table
missing_table <- data.frame(
  column = names(df_train),
  missing_count = sapply(df_train, function(x) sum(is.na(x))),
  missing_percent = sapply(df_train, function(x) mean(is.na(x)) * 100)
)

# we filter out the columns without any missing values and sort it in descending order
missing_table %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))


```

we see that `reviews_per_month` columns has 475 missing values - we try to examine it to see how we proceed (and check if N/A just means 0)

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train %>%
  filter(reviews_per_month == 0 | is.na(reviews_per_month))
```

we see that all rows that have N/A for `reviews_per_month` have `number_of_reviews` = 0 so we set `reviews_per_month` to zero and assigne listings without reviews with a dummy date outside the data range of our dataset: "2010-01-01"

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train <- df_train %>%
  mutate(
    reviews_per_month = tidyr::replace_na(reviews_per_month, 0),
    last_review = coalesce(last_review, as.Date("2010-01-01"))
  )

check_na <- function(df_train) {
  na_counts <- colSums(is.na(df_train))
  na_counts[na_counts > 0]
}

check_na(df_train)
```

### 1.4 - Transforming values

#### 1.4.1 - Creating ratios

basic and even advanced models such as neural networks and XGBoost have a hard time calculating ratios. hence in order to try to scale the variables as well as give them a ratio, we will turn

-   `availability_365_rate`: a percentage metric for the expected availability in a year `availability_365_rate` \<- (365-`availability_365`)/365

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train <- df_train %>%
  mutate(availability_365_rate=(365-availability_365)/365) %>%
  select(-availability_365)
```

-   we also do one-hot encoding of room_type, as it is less than 17 unique values.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

room_dummies <- model.matrix(~ room_type - 1, data = df_train) %>%
  as.data.frame()

# Bind the new dummy columns and remove the original column
df_train <- bind_cols(
  df_train %>% select(-room_type),
  room_dummies
)

```

#### 1.4.2 - Text-mining! :D

although the name serves as more of an identifier, it still contains a lot of adjectives describing the text, e.g here: Elegant vintage apartment in Ravenswood or phrases like this, which certainly don't make me feel so good about it: Disinfected, Sunny, Safe, top floor Condo w/ view. hence we could create columns counting the amount of words representing luxury terms, positive terms, and terms that are sketchy or show that is cheap

```{r, echo=TRUE, message=FALSE, warning=FALSE}
luxury_terms <- c(
  "luxury", "luxurious", "luxe", "penthouse", "chic", "elegant", 
  "opulent", "upscale", "high-end", "premium", "exclusive", "boutique", 
  "designer", "plush", "deluxe", "posh", "glamorous", "refined",
  "sophisticated", "loft", "mansion", "oasis", "stunning", "gorgeous", 
  "swanky", "spa", "suite", "modern", "sleek", "stylish", "iconic", 
  "duplex", "villa", "estate", "suite",
  "elite", "first-class", "grand", "gold", "coast,", "sanctuary", "haven", "oasis", "getaway", "hideaway", "king"
)

positive_terms <- c(
  "stunning","magnificent","spacious","great","cozy","beautiful","gorgeous", "dreamy",
  "amazing","lovely","bright","modern","sleek","stylish","sunny",
  "quiet","comfortable","renovated","brand[- ]new","updated","charming", "new"
)

sketchy_terms <- c(
  "safe", "clean", "sanitized",
  "tiny", "quirky", "basement", "subterranean", 
  "lower level", "garden unit", "pad", "nook", "crash pad", 
  "hostel", "price drop", "great deal", "bargain", "deal",
  "frugal", "cheap", "budget", "value",
  "unique", "artsy", "shared", "sofa", "couch", "futon", "loft"
)


# here we build a regex function to get all the terms from the list 
rx_or <- function(terms) paste0("(?i)\\b(", paste(terms, collapse="|"), ")\\b")

rx_luxury       <- rx_or(luxury_terms)
rx_positive     <- rx_or(positive_terms)
rx_sketchy  <- rx_or(sketchy_terms)

# # here we create a function to clean the name so that it's all lozercase, removing accents and also white spaces
normalize_name <- function(x) {
  x %>%
    str_trim() %>%
    str_squish() %>%
    str_to_lower()
}

# -- 4) Engineer features from `name` -----------------------------------------
df_train <- df_train %>%
  mutate(
    name_norm       = normalize_name(name),
    name_word_count = if_else(!is.na(name_norm) & name_norm != "",
                              str_count(name_norm, "\\S+"), 0L),
    name_special_char_rate = if_else(str_length(name) > 0,
                                     str_count(name, "[^\\p{L}\\p{N} .&]") / str_length(name),
                                     0),

    name_luxury_rate   = if_else(name_word_count > 0,
                                 str_count(name_norm, rx_luxury)   / name_word_count, 0),

    name_positive_rate = if_else(name_word_count > 0,
                                 str_count(name_norm, rx_positive) / name_word_count, 0),

    name_sketchy_rate  = if_else(name_word_count > 0,
                                 str_count(name_norm, rx_sketchy)  / name_word_count, 0),

    name_upper_word_rate = if_else(name_word_count > 0,
                                   str_count(name, "\\b\\p{Lu}+(?:'\\p{Lu}+)*\\b") / name_word_count, 0)
  ) %>%
  select(-name_norm, -name) # removing the redundant columns now

```

#### 1.4.3 - Augmenting with new data

We hypothesize that, as in most major cities, location is the main determinant of Airbnb prices in Chicago. To test this, we get new data based on the coordinates of each listing.\
The coordinates of Chicago’s central area were identified from public sources. Because Chicago’s downtown has a relatively rectangular shape, the definition of the city center can be considered reasonably objective.

For each listing, we calculated the distance from the city center (in kilometers). Additionally, using the "neighbourhood" variable to classify listings into downtown vs. non-downtown categories by marking the neighborhoods that belong to the central area.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

#plot of locations of listings
ggplot(df_train, aes( longitude, latitude, color = price )) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_viridis_c(trans = "log10") +
  labs(title = "Price (log) and locations")

# a new column "distance from center" in km
chicago_center <- c(-87.63, 41.88)
df_train$distance_from_center <- distHaversine(df_train[, c("longitude", "latitude")], chicago_center) / 1000  

# is the listing in the downtown? Y/N
# Caution: 'is_downtown' and 'distance_from_center' are collinear.
df_train <- df_train %>%
  mutate(
    is_downtown = if_else(
      neighbourhood %in% c("Loop", "Lower West Side", "Near West Side", "Near North Side", "Near South Side"),
      "1", "0"
    ) %>% as.factor()
  )
```

#### 1.4.4 - Log transformation

We check for right-skewed patterns in the numeric data, in order to make them more linear through log function.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

numeric_cols <- names(df_train)[sapply(df_train, is.numeric)]

numeric_plots <- lapply(numeric_cols, function(col) {
  ggplot(df_train, aes(x = .data[[col]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
    geom_density(color = "red", size = 1) +
    labs(title = paste("Distribution of", col), x = col, y = "Density") +
    theme_minimal()
})


for (p in numeric_plots) {
  print(p)
}

```

We see that the columns `number_of_reviews`, `reviews_per_month`, `calculated_host_listings_count` present a strong right skewed data, while the text mining rate columns present also a similar pattern. hence we pass the log function through them. `price` also presents the same, but it is the target variable so we leave it be

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train <- df_train %>%
  mutate(
    number_of_reviews = log1p(number_of_reviews),
    reviews_per_month = log1p(reviews_per_month),
    calculated_host_listings_count = log1p(calculated_host_listings_count),
    name_special_char_rate = log1p(name_special_char_rate),
    name_luxury_rate = log1p(name_luxury_rate),
    name_positive_rate = log1p(name_positive_rate),
    name_sketchy_rate = log1p(name_sketchy_rate),
    name_upper_word_rate = log1p(name_upper_word_rate),
  )

```

#### 1.4.5 - Outlier Removal

Now we remove outliers for the numeric columns which can technically have an infinite range (so pretty much anything aside from latitude, longitude, and rates)

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train <- df_train %>%
  filter(
    remove_outliers_iqr(minimum_nights),
    remove_outliers_iqr(number_of_reviews),
    remove_outliers_iqr(reviews_per_month),
    remove_outliers_iqr(calculated_host_listings_count),
    remove_outliers_iqr(name_word_count),
    remove_outliers_iqr(price)
  )

df_train

# Visualizations of the preprocessed data set
df_train %>%
  keep(is.numeric) %>%                     # Keep only numeric columns
  gather() %>%                             # Convert to key-value pairs
  ggplot(aes(value)) +                     # Plot the values
  facet_wrap(~ key, scales = "free") +   # In separate panels
  geom_histogram()                       # As histogram

# Visualizations of text features
text_feature_summary <- df_train %>%
  summarise(
    luxury_mean   = mean(name_luxury_rate, na.rm = TRUE),
    positive_mean = mean(name_positive_rate, na.rm = TRUE),
    sketchy_mean  = mean(name_sketchy_rate, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(),
               names_to = "feature",
               values_to = "mean_rate") %>%
  mutate(
    feature = recode(feature,
                     "luxury_mean"   = "Luxury terms",
                     "positive_mean" = "Positive terms",
                     "sketchy_mean"  = "Sketchy terms")
  )

# histograms of distributions
df_train %>%
  select(name_luxury_rate, name_positive_rate, name_sketchy_rate) %>%
  pivot_longer(cols = everything(),
               names_to = "feature",
               values_to = "rate") %>%
  mutate(feature = recode(feature,
                          "name_luxury_rate"   = "Luxury terms",
                          "name_positive_rate" = "Positive terms",
                          "name_sketchy_rate"  = "Sketchy terms")) %>%
  ggplot(aes(x = rate, fill = feature)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "black") +
  facet_wrap(~feature, scales = "free_y") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Distribution of Keyword Proportions in Airbnb Listing Titles",
    x = "Proportion of Matching Words in Title",
    y = "Number of Listings",
    fill = "Word Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# bar chart of average values
ggplot(text_feature_summary, aes(x = feature, y = mean_rate, fill = feature)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = percent(mean_rate, accuracy = 0.01)),
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Average Proportion of 'Sentiment' Words in Listing Titles",
    x = NULL, y = "Average Word Proportion"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

```

## Step 2. Data Analysis

### 2.1 Correlation Analysis

#### 2.1.1 Numerical Columns

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_train$last_review <- as.numeric(as.Date(df_train$last_review)) 
df_train$is_downtown <- as.numeric(as.character(df_train$is_downtown)) # we get the numeric value of last_review and is_downtown for the correlation analysis

numeric_cols <- names(df_train)[sapply(df_train, is.numeric)]

# vreate correlation matrix
cor_matrix <- cor(df_train[, numeric_cols], use = "complete.obs", method = "spearman")

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         number.cex = 0.45,   # the plot is too dense, so we're reducing the size of the texts
         tl.cex = 0.6
  )

# to make it more clear, we plot the correlations here for price in descending order 

cor_with_price <- sort(cor_matrix[ , "price"], decreasing = TRUE)

cor_table <- data.frame(
  variable = names(cor_with_price),
  spearman_correlation = cor_with_price,
  row.names = NULL
)

cor_table
```

The strongest positive correlation with price is room_type = Entire home/apt (+0.55), meaning full units are significantly more expensive. Luxury wording in the name (+0.18), minimum_nights (+0.13), and longitude (+0.19) show small positive effects, suggesting upscale wording, longer stays, or certain locations slightly increase price.

Weak positive correlations for features like name length or hotel room type indicate minimal influence. Special characters or uppercase words in the name have almost no impact.

On the negative side, distance_from_center (-0,32) room_type = Private room (-0.53) and Shared room (-0.15) strongly reduce price, as expected. More reviews and higher review frequency also slightly correlate with lower prices, possibly reflecting affordable, high-turnover listings. Other variables (availability, host listing count, positive/sketchy wording) show very small negative correlations.

Overall, room type and location is the dominant driver of price, while most other features have weak or negligible linear relationships.

#### 2.1.2 Categorical Columns

Spearman correlation does not work for categorical variables correctly (unless they are ordered and encoded correctly), hence we use kruskl-wallis test instead:

```{r, echo=TRUE, message=FALSE, warning=FALSE}

categorical_cols <- names(df_train)[sapply(df_train, function(x) is.factor(x) || is.character(x))]

cat_results <- data.frame(
  variable = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (col in categorical_cols) {
  temp_df <- df_train[!is.na(df_train[[col]]) & !is.na(df_train$price), ]
  if (length(unique(temp_df[[col]])) > 1) {
    test <- kruskal.test(price ~ temp_df[[col]], data = temp_df)
    cat_results <- rbind(cat_results, data.frame(variable = col, p_value = test$p.value))
  }
}

cat_results <- cat_results[order(cat_results$p_value), ]

print(cat_results)

df_train

```

The p-value is EXTREMELY low, meaning that the neighborhood is statistically significant for the prediction of price.

## Step 3. Modelling

Our correlation analysis has displayed that the room type and location seem to have a statistically significant correlation with the price. Also the rate in which luxury terms appear, the minimum nights per booking seem to have a strong effect.

For this modelling step, we decided to proceed with XGBoost, as it deals N/A values in production + also comes with built-in regularization and efficient use of random forest. Also we can use it to determine the importance of the variables in the decision making.

Before that we made a quick check of other possible models to make sure that XGBoost is tha best option on our data, as we assume.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Test other models to make sure, that XGBoost model gives the best outcomes.

set.seed(42)

# 1. Preparation
y <- "price"

rec <- recipe(price ~ ., data = df_train) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

folds <- vfold_cv(df_train, v = 5)
metrics <- yardstick::metric_set(yardstick::rmse)

# 2. Models
# OLS
ols_spec <- linear_reg(mode = "regression") %>%
  set_engine("lm")

# Ridge
ridge_spec <- linear_reg(
  mode = "regression",
  penalty = tune(),
  mixture = 0   # mixture=0 → Ridge
) %>% set_engine("glmnet")

# Lasso
lasso_spec <- linear_reg(
  mode = "regression",
  penalty = tune(),
  mixture = 1   # mixture=1 → Lasso
) %>% set_engine("glmnet")

# Random Forest
rf_spec <- rand_forest(
  mode = "regression",
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>% set_engine("ranger")

# Support Vector Regression
svr_spec <- svm_rbf(
  mode = "regression",
  cost = tune(),
  rbf_sigma = tune()
) %>% set_engine("kernlab")

# 3. workflows
wf_ols <- workflow() %>% add_recipe(rec) %>% add_model(ols_spec)
wf_ridge <- workflow() %>% add_recipe(rec) %>% add_model(ridge_spec)
wf_lasso <- workflow() %>% add_recipe(rec) %>% add_model(lasso_spec)
wf_rf <- workflow() %>% add_recipe(rec) %>% add_model(rf_spec)
wf_svr <- workflow() %>% add_recipe(rec) %>% add_model(svr_spec)

# 4. Nets for hyperparameters
ridge_grid <- grid_regular(penalty(range = c(-4, 1)), levels = 10)
lasso_grid <- grid_regular(penalty(range = c(-4, 1)), levels = 10)

rf_grid <- grid_latin_hypercube(
  mtry(range = c(5, 20)),
  min_n(range = c(2, 10)),
  size = 10
)

svr_grid <- grid_latin_hypercube(
  cost(range = c(-2, 2)),
  rbf_sigma(range = c(-2, 0)),
  size = 10
)

ctrl <- control_grid(verbose = TRUE)

# 5. Cross-validation
res_ols <- fit_resamples(wf_ols, resamples = folds, metrics = metrics, control = ctrl)
res_ridge <- tune_grid(wf_ridge, resamples = folds, grid = ridge_grid, metrics = metrics, control = ctrl)
res_lasso <- tune_grid(wf_lasso, resamples = folds, grid = lasso_grid, metrics = metrics, control = ctrl)
res_rf <- tune_grid(wf_rf, resamples = folds, grid = rf_grid, metrics = metrics, control = ctrl)
res_svr <- tune_grid(wf_svr, resamples = folds, grid = svr_grid, metrics = metrics, control = ctrl)

# 6. Comparision
collect_metrics(res_ols) %>% mutate(model="OLS")
collect_metrics(res_ridge) %>% mutate(model="Ridge")
collect_metrics(res_lasso) %>% mutate(model="Lasso")
collect_metrics(res_rf) %>% mutate(model="Random Forest")
collect_metrics(res_svr) %>% mutate(model="SVR") -> all_metrics

all_metrics_df <- bind_rows(
  collect_metrics(res_ols) %>% mutate(model="OLS"),
  collect_metrics(res_ridge) %>% mutate(model="Ridge"),
  collect_metrics(res_lasso) %>% mutate(model="Lasso"),
  collect_metrics(res_rf) %>% mutate(model="Random Forest"),
  collect_metrics(res_svr) %>% mutate(model="SVR")
)

# RMSE 
all_metrics_df %>%
  filter(.metric == "rmse") %>%
  select(model, mean, std_err) %>%
  arrange(mean)
```

The best result has Random Forest with RSME 67.52. We just leave it here for now and get back to it later.

### 3.1 Hyperparameter turning

We proceed by tuning the XGBoost parameters, having RMSE for the loss function.

First we split into training and validation in order to ensure the hyperparameters are not overfitting the data. There's a lot of data, hence we proceed with a 80/20 split

```{r, echo=TRUE, message=FALSE, warning=FALSE}

data_split <- initial_split(df_train, prop = 0.8)
train_df   <- training(data_split)
val_df    <- testing(data_split)
y = "price"
p <- ncol(dplyr::select(train_df, -dplyr::all_of(y)))
mtry_fixed <- max(1L, floor(0.8 * p))

rec <- recipe(price ~ ., data = train_df) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

```

Then we proceed with the setting of the hyperparametes to tune - there are many to count, and in general we just decided to go with the best practices outlined on XGBoost: <https://www.analyticsvidhya.com/blog/2016/03/complete-guide-parameter-tuning-xgboost-with-codes-python/>

#### 3.1.1 Step 1: Selection of Learning Rate and Number of Estimators for Tuning Tree-Based Parameters.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

set.seed(42) # setting seed here once again just to be safe :D

#Creating cross-validation and metrics
folds   <- vfold_cv(train_df, v = 5)
metrics <- yardstick::metric_set(yardstick::rmse)
ctrl    <- control_grid(verbose = TRUE)
wf_from_spec <- function(spec) workflow() %>% add_recipe(rec) %>% add_model(spec)

#Setting the XGBoost model
mtry_fixed <- max(1L, floor(0.8 * p))

#Configuring the XGBoost model
spec_step1 <- boost_tree(
  mode           = "regression",
  trees          = tune(),
  tree_depth     = 5L,
  learn_rate     = tune(),
  min_n          = 1L,
  loss_reduction = 0.15,
  mtry           = mtry_fixed,
  sample_size    = 0.8
) %>%
  set_engine("xgboost", eval_metric = "rmse",
             nthread = max(1, parallel::detectCores() - 1))

#Defining the workflow build function
wf_step1 <- wf_from_spec(spec_step1)

# A sensible LR set + broad trees range. We form hyperparameter grid
rates <- c(0.3, 0.2, 0.1, 0.05, 0.03, 0.025, 0.01, 0.005)
grid_step1 <- tidyr::expand_grid(
  learn_rate = rates,
  trees      = c(5L, 10L, 25L, 50L, 75L, 100L, 150L, 200L, 400, 500L, 750L, 1000L, 1200L)
)

#Starting the selection of hyperparameters
res_step1  <- tune_grid(wf_step1, resamples = folds, grid = grid_step1,
                        metrics = metrics, control = ctrl)

best_step1 <- select_best(res_step1, metric = "rmse")

best_trees <- best_step1$trees
best_lr    <- best_step1$learn_rate

#Generating a performance table
perf_step1 <- make_perf_table(res_step1, wf_step1, train_df, truth = y)
perf_step1 %>% slice_head(n = 15)
perf_step1

```

The line with "trees 750, learn_rate 0.010, RSME 65.23719, gap 13.53" looks like the best combination of RSME and gap.

#### 3.1.2 Step 2: Tune max_depth and min_child_weight

```{r, echo=TRUE, message=FALSE, warning=FALSE}

#Defining the model for the 2. stage
spec_step2 <- boost_tree(
  mode           = "regression",
  trees          = best_trees,
  tree_depth     = tune(),
  min_n          = tune(),
  learn_rate = best_lr,
  loss_reduction = 0.15,
  mtry           = mtry_fixed,
  sample_size    = 0.8
) %>% set_engine("xgboost", eval_metric = "rmse",
                 nthread = max(1, parallel::detectCores() - 1))

#Creating a workflow
wf_step2 <- wf_from_spec(spec_step2)

#Creating a hyperparameter space
grid_step2 <- grid_latin_hypercube(
  tree_depth(c(3L, 12L)),
  min_n(c(1L, 50L)),
  size = 20
)

#CV
res_step2  <- tune_grid(wf_step2, resamples = folds,
                        grid = grid_step2, metrics = metrics, control = ctrl)

#Choosing the best combination
best_step2 <- select_best(res_step2, metric = "rmse")

perf_step2 <- make_perf_table(res_step2, wf_step2, train_df, truth = y)
perf_step2 %>% slice_head(n = 15)
perf_step2
```

#### 3.1.3 Step 3: Tune gamma

```{r, echo=TRUE, message=FALSE, warning=FALSE}

spec_step3 <- boost_tree(
  mode           = "regression",
  trees          = best_trees,
  tree_depth     = best_step2$tree_depth,
  min_n          = best_step2$min_n,
  learn_rate     = best_lr,
  loss_reduction = tune(),
  mtry           = mtry_fixed,
  sample_size    = 0.8
) %>% set_engine("xgboost", eval_metric = "rmse",
                 nthread = max(1, parallel::detectCores() - 1))

wf_step3 <- wf_from_spec(spec_step3)

grid_step3 <- grid_latin_hypercube(
  loss_reduction(c(-4, 2)),   # log10 range ≈ [1e-4, 1e2]
  size = 25
)


res_step3  <- tune_grid(wf_step3, resamples = folds,
                        grid = grid_step3, metrics = metrics, control = ctrl)
best_step3 <- select_best(res_step3, metric = "rmse")

gamma_val <- best_step3$loss_reduction
if (is.numeric(gamma_val) && gamma_val < 0) gamma_val <- 10^gamma_val

perf_step3 <- make_perf_table(res_step3, wf_step3, train_df, truth = y)
perf_step3 %>% slice_head(n = 15)
perf_step3

```

#### 3.1.4 Step 4: Tune subsample and colsample_bytree

```{r, echo=TRUE, message=FALSE, warning=FALSE}

props <- seq(0.5, 0.9, by = 0.1)
grid_step4 <- tidyr::expand_grid(
  sample_size = props,
  mtry        = unique(as.integer(round(seq(0.5, 0.9, by = 0.1) * p)))
)


spec_step4 <- boost_tree(
  mode           = "regression",
  trees          = best_trees,
  tree_depth     = best_step2$tree_depth,
  min_n          = best_step2$min_n,
  learn_rate     = best_lr,
  loss_reduction = {               
    g <- best_step3$loss_reduction
    if (is.numeric(g) && g < 0) 10^g else g
  },
  sample_size    = tune(),
  mtry           = tune()
) %>% set_engine(
  "xgboost",
  eval_metric = "rmse",
  nthread     = max(1, parallel::detectCores() - 1)
)

wf_step4  <- wf_from_spec(spec_step4)
res_step4 <- tune_grid(
  wf_step4,
  resamples = folds,
  grid      = grid_step4,
  metrics   = metrics,
  control   = ctrl
)

best_step4 <- select_best(res_step4, metric = "rmse")

perf_step4 <- make_perf_table(res_step4, wf_step4, train_df, truth = y)
perf_step4 %>% slice_head(n = 15)
perf_step4

```

#### 3.1.5 Step 5: Tuning Regularization Parameters

```{r, echo=TRUE, message=FALSE, warning=FALSE}

grid_step5 <- tibble::tibble(
  lambda = 10^runif(25, -6, 3),  # 1e-6 .. 1e3
  alpha  = 10^runif(25, -8, 2)   # 1e-8 .. 1e2
)

spec_step5 <- boost_tree(
  mode           = "regression",
  trees          = best_trees,
  tree_depth     = best_step2$tree_depth,
  learn_rate     = best_lr,
  min_n          = best_step2$min_n,
  loss_reduction = { g <- best_step3$loss_reduction; if (is.numeric(g) && g < 0) 10^g else g },
  mtry           = best_step4$mtry,
  sample_size    = best_step4$sample_size
) %>% set_engine(
  "xgboost",
  eval_metric = "rmse",
  nthread     = max(1, parallel::detectCores() - 1),
  lambda      = tune(),
  alpha       = tune()
)

wf_step5  <- wf_from_spec(spec_step5)
res_step5 <- tune_grid(wf_step5, resamples = folds, grid = grid_step5, metrics = metrics, control = ctrl)
best_step5 <- select_best(res_step5, metric = "rmse")

perf_step5 <- make_perf_table(res_step5, wf_step5, train_df, truth = y)
perf_step5 %>% slice_head(n = 15)
perf_step5

```

### Step 4. Evaluation of Model Tuning Process

```{r, echo=TRUE, message=FALSE, warning=FALSE}

grab_step <- function(res, step, fixed = list(), summarize = TRUE) {
  out <- collect_metrics(res, summarize = summarize) %>%
    filter(.metric == "rmse") %>%
    mutate(step = step)
  if (length(fixed)) {
    out <- bind_cols(out, as_tibble(fixed))
  }
  out
}

param_cols <- c("trees","learn_rate","tree_depth","min_n","loss_reduction",
                "sample_size","mtry","lambda","alpha")

step1_tbl <- grab_step(
  res_step1, "step1_trees_lr",
  fixed = list(
    tree_depth = 5L,
    min_n = 1L,
    loss_reduction = 0.15,
    mtry = mtry_fixed,
    sample_size = 0.8
  )
)

step2_tbl <- grab_step(
  res_step2, "step2_depth_min_n",
  fixed = list(
    trees = best_trees,
    learn_rate = best_lr,
    loss_reduction = 0.15,
    mtry = mtry_fixed,
    sample_size = 0.8
  )
)

step3_tbl <- grab_step(
  res_step3, "step3_gamma",
  fixed = list(
    trees = best_trees,
    learn_rate = best_lr,
    tree_depth = best_step2$tree_depth,
    min_n = best_step2$min_n,
    mtry = mtry_fixed,
    sample_size = 0.8
  )
) %>%
  mutate(loss_reduction_effective =
           ifelse(is.numeric(loss_reduction) & loss_reduction < 0,
                  10^loss_reduction, loss_reduction))

step4_tbl <- grab_step(
  res_step4, "step4_subsample_mtry",
  fixed = list(
    trees = best_trees,
    learn_rate = best_lr,
    tree_depth = best_step2$tree_depth,
    min_n = best_step2$min_n,
    loss_reduction = { g <- best_step3$loss_reduction; if (is.numeric(g) && g < 0) 10^g else g }
  )
)

step5_tbl <- grab_step(
  res_step5, "step5_lambda_alpha",
  fixed = list(
    trees = best_trees,
    learn_rate = best_lr,
    tree_depth = best_step2$tree_depth,
    min_n = best_step2$min_n,
    loss_reduction = { g <- best_step3$loss_reduction; if (is.numeric(g) && g < 0) 10^g else g },
    mtry = best_step4$mtry,
    sample_size = best_step4$sample_size
  )
)

all_results <- bind_rows(step1_tbl, step2_tbl, step3_tbl, step4_tbl, step5_tbl) %>%
  select(step, any_of(c("trees","learn_rate","tree_depth","min_n",
                        "loss_reduction","sample_size","mtry","lambda","alpha",
                        "mean","n","std_err",".config"))) %>%
  mutate(
    lambda = dplyr::coalesce(lambda, 1),
    alpha  = dplyr::coalesce(alpha, 0)
  ) %>%
  arrange(mean)

all_results %>% slice_head(n = 20)

```

It is found from the examination of the table that the the hyperparameter optimization does not yield much difference between the results of Steps 1-5 and overall evaluation of Model Tuning Process, and the improval is quite minimal. Thus, we find one best model, with the parameters: trees: 1200, learn_rate: 0.1, tree_depth: 6, min_n: 10, loss_reduction: 0.0245, sample: 0.9, mtry: 13 lambda: 1.0 (default) alpha: 0.0 (default).

### Step 5. Deployment

here we use our best model to infer the data.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

df_test = read.csv("data/test.csv")

preprocess_abnb <- function(dat) {
  dat <- dat %>%
    # drop pure identifiers, cast types
    select(-ID, -host_id, -host_name) %>%
    mutate(
      neighbourhood = as.factor(neighbourhood),
      last_review = ymd(last_review),
      last_review = coalesce(last_review, as.Date("2010-01-01")),
      last_review = as.numeric(last_review),
      distance_from_center = distHaversine(dat[, c("longitude", "latitude")], chicago_center) / 1000,
      is_downtown = if_else(neighbourhood %in% c("Loop", "Lower West Side", "Near West Side", "Near North Side", "Near South Side"),
      1, 0  ), 
      reviews_per_month = tidyr::replace_na(reviews_per_month, 0)
  )
    
  dat <- dat %>%
    mutate(availability_365_rate = (365 - availability_365) / 365) %>%
    select(-availability_365)

  room_dummies <- model.matrix(~ room_type - 1, data = dat) %>% as.data.frame()
  dat <- bind_cols(dat %>% select(-room_type), room_dummies)

  train_dummy_cols <- grep("^room_type", names(df_train), value = TRUE)
  missing_cols <- setdiff(train_dummy_cols, names(dat))
  if (length(missing_cols)) {
    dat[missing_cols] <- 0L
  }

  dat <- dat %>%
    mutate(
      name_norm = normalize_name(name),
      name_word_count = if_else(!is.na(name_norm) & name_norm != "",
                                str_count(name_norm, "\\S+"), 0L),
      name_special_char_rate = if_else(str_length(name) > 0,
                                       str_count(name, "[^\\p{L}\\p{N} .&]") / str_length(name), 0),
      name_luxury_rate   = if_else(name_word_count > 0,
                                   str_count(name_norm, rx_luxury)   / name_word_count, 0),
      name_positive_rate = if_else(name_word_count > 0,
                                   str_count(name_norm, rx_positive) / name_word_count, 0),
      name_sketchy_rate  = if_else(name_word_count > 0,
                                   str_count(name_norm, rx_sketchy)  / name_word_count, 0),
      name_upper_word_rate = if_else(name_word_count > 0,
                                     str_count(name, "\\b\\p{Lu}+(?:'\\p{Lu}+)*\\b") / name_word_count, 0)
    ) %>%
    select(-name_norm, -name)

  dat <- dat %>%
    mutate(
      number_of_reviews = log1p(number_of_reviews),
      reviews_per_month = log1p(reviews_per_month),
      calculated_host_listings_count = log1p(calculated_host_listings_count),
      name_special_char_rate = log1p(name_special_char_rate),
      name_luxury_rate = log1p(name_luxury_rate),
      name_positive_rate = log1p(name_positive_rate),
      name_sketchy_rate = log1p(name_sketchy_rate),
      name_upper_word_rate = log1p(name_upper_word_rate)
    )

  dat
}

test_ids <- df_test$ID

df_test_proc <- preprocess_abnb(df_test)

final_rec <- recipe(price ~ ., data = df_train) %>%
step_zv(all_predictors()) %>%
step_impute_median(all_numeric_predictors()) %>%
step_other(all_nominal_predictors(), threshold = 0.01) %>%
step_novel(all_nominal_predictors()) %>%
step_dummy(all_nominal_predictors())

p_final      <- ncol(dplyr::select(df_train, -price))
mtry_chosen  <- min(47L, p_final)  

final_spec <- boost_tree(
mode           = "regression",
trees          = 1200L,
learn_rate     = 0.01,
tree_depth     = 6L,
min_n          = 10L,
loss_reduction = 0.0245,  # gamma
sample_size    = 0.9,
mtry           = mtry_chosen
) %>%
set_engine(
"xgboost",
eval_metric = "rmse",
lambda      = 9.4e-6,
alpha       = 1.7e-8,
nthread     = max(1, parallel::detectCores() - 1)
)

final_wf <- workflow() %>%
add_recipe(final_rec) %>%
add_model(final_spec)

final_fit <- fit(final_wf, data = df_train)

test_preds <- predict(final_fit, new_data = df_test_proc) %>%
dplyr::rename(price = .pred) %>%
dplyr::mutate(price = pmax(price, 0)) 

submission <- tibble::tibble(
ID    = test_ids,
price = test_preds$price
)

readr::write_csv(submission, "predictions_submission.csv")

head(submission, 10)

data_split <- initial_split(df_train, prop = 0.8)
train_df   <- training(data_split)
val_df     <- testing(data_split)

rec_prep <- prep(final_rec)

train_prep <- bake(rec_prep, new_data = train_df)
val_prep   <- bake(rec_prep, new_data = val_df)

model_train_only <- final_spec %>%
  fit(price ~ ., data = train_prep)

train_preds_split <- predict(model_train_only, new_data = train_prep) %>% pull(.pred)
val_preds_split   <- predict(model_train_only, new_data = val_prep)   %>% pull(.pred)

train_rmse_split <- rmse_vec(truth = train_prep$price, estimate = train_preds_split)
val_rmse_split   <- rmse_vec(truth = val_prep$price, estimate = val_preds_split)

cat(sprintf("Train RMSE (split model): %.4f\n", train_rmse_split))
cat(sprintf("Validation RMSE: %.4f\n", val_rmse_split))
```
